Analysis: Exploratory Data Analysis

### 2021-12-13 code/analysis_code/analysis-eda.Rmd
Analysis Complete

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
* skimr: for data summarization
* summarytools: for overall dataframe summary
* ggplot2: for plotting data
* car: for creating QQ plots
* gtsummary: for creating tables for summary statistics / other numerical outputs
* scales: for percent calculation and axis formatting
* cowplot: for placing plots side by side
* table1: for making summary tables

@@

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
library(skimr) #for data summarization
library(summarytools) #for overall dataframe summary
library(ggplot2) #for plotting data
library(car) #for creating QQ plots
library(gtsummary) #for creating tables
library(scales) #for percent calculation
library(cowplot) #for placing plots side by side
library(table1) #for making summary tables

@@

@@

## Outcome of Interest: Requested FEMA Funds
The first of two primary outcomes of interest is the total amount of funds the state requested from the federal government for a disaster declaration.
```{r}
#summary statistics
base::summary(EDAdata$ReqAmt)

#density plot
ggplot2::ggplot(data = EDAdata, aes(x = ReqAmt)) +
  geom_density()

#try a log transformation out of curiosity 
ggplot2::ggplot(data = EDAdata, aes(x = log(ReqAmt))) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
  geom_boxplot()

#crop to $1M to get a better idea
ggplot2::ggplot(data = EDAdata, aes(y = ReqAmt)) +
  geom_boxplot() +
  ylim(0, 1000000)

#log transformed boxplot
ggplot2::ggplot(data = EDAdata, aes(y = log(ReqAmt))) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata$ReqAmt)
```
These outputs show the requested amounts do not follow a normal distribution, which is to be expected given the broad range of disasters. We may need to remove outliers in the analysis, but for now, we'll leave them to capture the true nature of disaster declarations. The largest outlier is likely Hurricane Maria. The log transformation does a better job of showing the distribution of the data, so we may need to consider an exponential model in the analysis. The transformed and untransformed figures suggest there are significant tails in the data.

@@

## Outcome of Interest: Obligated FEMA Funds
The alternative outcome of interest is the funding obligated from FEMA. While this may be the same as `ReqAmt` (especially for large-scale disasters), FEMA doesn't always pay the amount the states request.
```{r obligated funds}
#summary statistics
base::summary(EDAdata$OblAmt)

#density plot
ggplot2::ggplot(data = EDAdata, aes(x = OblAmt)) +
  geom_density()

#try a log transformation out of curiosity 
ggplot2::ggplot(data = EDAdata, aes(x = log(OblAmt))) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata, aes(y = OblAmt)) +
  geom_boxplot()

#crop to $1M to get a better idea
ggplot2::ggplot(data = EDAdata, aes(y = OblAmt)) +
  geom_boxplot() +
  ylim(0, 1000000)

#log transformed boxplot
ggplot2::ggplot(data = EDAdata, aes(y = log(OblAmt))) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata$OblAmt)
```
It appears that obligated funds and requested funds have similar distributions and characteristics. The same commentary about the requested funds analysis applies here as well.

Before moving on, I want to create a plot of Requested vs obligated Funds to understand the relationship between the two outcomes:
```{r}
funds_comp <- EDAdata %>%
                ggplot2::ggplot(aes(x = log(ReqAmt), y = log(OblAmt))) +
                         geom_point()

funds_comp
```

@@

## Predictor: State
```{r}
#frequency table since it's categorical
#hide NAs because there are no NAs
summarytools::freq(EDAdata$state, report.nas = FALSE)

#examine graphical relationship with outcomes
#include a jitter function to have a better idea of number of measurements and distribution
#filter top 5 states
#start with obligated amount
obl_fig_state <- EDAdata %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
                    ggplot2::ggplot(aes(x = state, y = OblAmt)) +
                      geom_boxplot() +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      labs(x = "State or Territory",
                           y = "Obligated FEMA Funds",
                           title = "FEMA funds obligated to five most disaster-prone states")
obl_fig_state

#save file
obl_state_fig_file = here("results","obl-state.png")
ggsave(filename = obl_state_fig_file, plot = obl_fig_state)

#repeat for requested funds
req_fig_state <- EDAdata %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
                    ggplot2::ggplot(aes(x = state, y = ReqAmt)) +
                      geom_boxplot() +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      labs(x = "State or Territory",
                           y = "Requested FEMA Funds",
                           title = "FEMA funds requested by five most disaster-prone states")
req_fig_state

#save file
req_state_fig_file = here("results","req-state.png")
ggsave(filename = req_state_fig_file, plot = req_fig_state)

#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
decs_state_5 <- EDAdata %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dense_rank(-n) < 6) %>%
                    gtsummary::select(state, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = state,
                      label = list(state ~ "State or Territory",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**State**") %>%
                    gtsummary::modify_caption("**Table 4. FEMA Funding for 5 Most Disaster-Prone States 2012 - 2021**") %>%
                    gtsummary::bold_labels()
decs_state_5

@@

## Predictor: Incident Year
```{r}
#summary statistics since it's continuous
base::summary(EDAdata$IncidentYear)

@@

#density plot
ggplot2::ggplot(data = EDAdata, aes(x = IncidentYear)) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata, aes(y = IncidentYear)) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata$IncidentYear)

#histogram distribution
year_hist <- EDAdata %>% 
               ggplot2::ggplot(aes(x=IncidentYear)) + 
                        geom_bar(fill="steelblue4") + 
                        scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
                        scale_y_continuous(expand = c(0, 0), limits = c(0, 125)) +
                        geom_text(stat='count', aes(label=..count..), vjust=-1) +
                        labs(x = "Incident Year",
                             y = "Number of Declarations",
                             title = "Disaster Declarations per Year (2012 - 2021)")
year_hist

@@

#examine graphical relationship with outcomes of interest
#look at total cost per incident year
req_fig_year <- EDAdata %>%
                    ggplot2::ggplot(aes(x = IncidentYear)) +
                      geom_bar(aes(y = ReqAmt),
                               stat = "identity",
                               fill = "tomato") +
                      scale_y_continuous(expand = c(0, 0), 
                                         labels = scales::dollar_format()) +
                      scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
                      labs(x = "Incident Year",
                           y = "Funds ($)",
                           subtitle = "Requested")
req_fig_year

#repeat for obligated funds
obl_fig_year <- EDAdata %>%
                    ggplot2::ggplot(aes(x = IncidentYear)) +
                      geom_bar(aes(y = OblAmt),
                               stat = "identity",
                               fill = "turquoise4") +
                      scale_y_continuous(expand = c(0, 0),
                                         labels = scales::dollar_format()) +
                      scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
                      labs(x = "Incident Year",
                           y = "Funds ($)",
                           subtitle = "Obligated")
obl_fig_year

@@

title1 <- cowplot::ggdraw() + 
                    draw_label(
                      "Requested vs obligated FEMA funds per year",
                      fontface = 'bold',
                      x = 0,
                      hjust = 0) +
                    theme( plot.margin = margin(0, 0, 0, 7))

@@

#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
decs_year_3 <- EDAdata %>%
                    dplyr::add_count(IncidentYear) %>%
                    dplyr::filter(dense_rank(-n) < 4) %>%
                    gtsummary::select(IncidentYear, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = IncidentYear,
                      label = list(IncidentYear ~ "Year",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Year**") %>%
                    gtsummary::modify_caption("**Table 2. FEMA Funding for the 3 Costliest Years (2012 - 2021)**") %>%
                    gtsummary::bold_labels()
decs_year_3

@@

## Incident Month
```{r}
#summary statistics since it's continuous
base::summary(EDAdata$IncidentMonth)

#density plot
ggplot2::ggplot(data = EDAdata, aes(x = IncidentMonth)) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata, aes(y = IncidentMonth)) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata$IncidentMonth)

#histogram distribution
month_hist <- EDAdata %>% 
               ggplot2::ggplot(aes(x=IncidentMonth)) + 
                        geom_bar(fill="darkseagreen4") + 
                        scale_x_continuous(breaks = seq(1, 12, by = 1),
                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
                        scale_y_continuous(expand = c(0, 0), limits = c(0, 110)) +
                        geom_text(stat='count', aes(label=..count..), vjust=-1) +
                        labs(x = "Declaration Month",
                             y = "Number of Declarations",
                             title = "Disaster Declarations per Month (2012 - 2021)")
month_hist

#save file
month_fig_file = here("results","incidentmonth.png")
ggsave(filename = month_fig_file, plot = month_hist)

#examine graphical relationship with outcomes of interest
#look at total cost per incident month
req_fig_month <- EDAdata %>%
                    ggplot2::ggplot(aes(x = IncidentMonth)) +
                      geom_bar(aes(y = ReqAmt),
                               stat = "identity",
                               fill = "tomato") +
                      scale_y_continuous(expand = c(0, 0), 
                                         labels = scales::dollar_format()) +
                      scale_x_continuous(breaks = seq(1, 12, by = 1),
                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
                      labs(x = "Declaration Month",
                           y = "Funds ($)",
                           subtitle = "Requested")
req_fig_month

#repeat for obligated funds
obl_fig_month <- EDAdata %>%
                    ggplot2::ggplot(aes(x = IncidentMonth)) +
                      geom_bar(aes(y = OblAmt),
                               stat = "identity",
                               fill = "turquoise4") +
                      scale_y_continuous(expand = c(0, 0),
                                         labels = scales::dollar_format()) +
                      scale_x_continuous(breaks = seq(1, 12, by = 1),
                                           labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                                                      "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")) +
                      labs(x = "Declaration Month",
                           y = "Funds ($)",
                           subtitle = "Obligated")
obl_fig_month

#combine into one plot
plot_col1 <- cowplot::plot_grid(req_fig_month, 
                                     obl_fig_month,
                                     ncol = 1,
                                     align = "v",
                                     rel_heights = c(0.5,0.5))
plot_col1

title2 <- cowplot::ggdraw() + 
                    draw_label(
                      "Requested vs obligated FEMA funds by declaration month",
                      fontface = 'bold',
                      x = 0,
                      hjust = 0) +
                    theme( plot.margin = margin(0, 0, 0, 7))

funds_fig_month <- cowplot::plot_grid(
                              title2, plot_col1,
                              ncol = 1,
                              # rel_heights values control vertical title margins
                              rel_heights = c(0.1, 1))
funds_fig_month

#save file
funds_month_fig_file = here("results","funds-month.png")
ggsave(filename = funds_month_fig_file, plot = funds_fig_month)

#create a table where columns represent requested and obligated funds and rows are 3 most distaster prone-months
decs_month_3 <- EDAdata %>%
                    dplyr::add_count(IncidentMonth) %>%
                    dplyr::filter(dense_rank(-n) < 4) %>%
                    gtsummary::select(IncidentMonth, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = IncidentMonth,
                      label = list(IncidentMonth ~ "Month",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>%
                    gtsummary::modify_header(update = list(
                                               stat_1 ~ "**Jan**",
                                               stat_2 ~ "**Sept**",
                                               stat_3 ~ "**Oct**"
                                             )) %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Month**") %>%
                    gtsummary::modify_caption("**Table 3. FEMA Funding for the 3 Costliest Months (2012 - 2021)**") %>%
                    gtsummary::bold_labels()
decs_month_3

#save file
decs_month_file = here("results", "decs-month-3.Rds")
saveRDS(decs_month_3, file = decs_month_file)
```

@@

## Predictor: Declaration Type
```{r}
#frequency table since it's categorical
#hide NAs because there are no NAs
summarytools::freq(EDAdata$declarationType, report.nas = FALSE)

#examine graphical relationship with outcomes
#include a jitter function to have a better idea of number of measurements and distribution
#start with obligated amount
obl_fig_DT <- EDAdata %>%
                    ggplot2::ggplot(aes(x = declarationType, y = OblAmt)) +
                      geom_boxplot() +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "tomato") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      scale_x_discrete(labels = c("Major Disaster Declaration", "Emergency Declaration")) + 
                      labs(x = "Declaration Type",
                           y = "Obligated FEMA Funds",
                           title = "FEMA funds obligated for declaration types")
obl_fig_DT

#save file
obl_DT_fig_file = here("results","obl-DT.png")
ggsave(filename = obl_DT_fig_file, plot = obl_fig_DT)

#repeat for requested funds
req_fig_DT <- EDAdata %>%
                    ggplot2::ggplot(aes(x = declarationType, y = ReqAmt)) +
                      geom_boxplot() +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "turquoise4") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      scale_x_discrete(labels = c("Major Disaster Declaration", "Emergency Declaration")) + 
                      labs(x = "Declaration Type",
                           y = "Requested FEMA Funds",
                           title = "FEMA funds requested for declaration types")
req_fig_DT

#save file
req_DT_fig_file = here("results","req-DT.png")
ggsave(filename = req_DT_fig_file, plot = req_fig_DT)

#create a table where rows represent requested and obligated funds and columns are declaration type
decs_dtype <- EDAdata %>%
                    gtsummary::select(declarationType, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = declarationType,
                      label = list(declarationType ~ "Type",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_header(update = list(
                                               stat_1 ~ "**Major Disaster**",
                                               stat_2 ~ "**Emergency**")) %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2") ~ "**Type**") %>%
                    gtsummary::modify_caption("**Table 4. FEMA Funding for Declaration Types 2012 - 2021**") %>%
                    gtsummary::bold_labels()
decs_dtype

>>>>

* here: for path setting
* tidyverse: for all packages in the Tidyverse (ggplot2, dyplr, tidyr, readr, purr, tibble, stringr, forcats)
* skimr: for data summarization
* summarytools: for overall dataframe summary
* ggplot2: for plotting data
* car: for creating QQ plots
* gtsummary: for creating tables for summary statistics / other numerical outputs
* scales: for percent calculation and axis formatting
* cowplot: for placing plots side by side
* table1: for making summary tables
* maps: for making maps

@@

#load required packages
library(here) #to set paths
library(tidyverse) #for data processing
library(skimr) #for data summarization
library(summarytools) #for overall dataframe summary
library(ggplot2) #for plotting data
library(car) #for creating QQ plots
library(gtsummary) #for creating tables
library(scales) #for percent calculation
library(cowplot) #for placing plots side by side
library(table1) #for making summary tables
library(maps) #for making maps

@@

Before moving through the entire EDA, I want to create a secondary dataset with the outliers removed. This is the data actually used in the ML analysis, so this is the dataset that should be used for the EDA.
```{r outliers}
EDAdata_outliers <- EDAdata %>%
                      dplyr::filter(log(ReqAmt) > 0)
```

@@

## Outcome of Interest: Requested FEMA Funds
The first of two primary outcomes of interest is the total amount of funds the state requested from the federal government for a disaster declaration.
```{r}
#summary statistics
base::summary(EDAdata_outliers$ReqAmt)

#density plot
ggplot2::ggplot(data = EDAdata_outliers, aes(x = ReqAmt)) +
  geom_density()

#try a log transformation out of curiosity 
ggplot2::ggplot(data = EDAdata_outliers, aes(x = log(ReqAmt))) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata_outliers, aes(y = ReqAmt)) +
  geom_boxplot()

#crop to $1M to get a better idea
ggplot2::ggplot(data = EDAdata_outliers, aes(y = ReqAmt)) +
  geom_boxplot()

#log transformed boxplot
ggplot2::ggplot(data = EDAdata_outliers, aes(y = log(ReqAmt))) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata_outliers$ReqAmt)
```
These outputs show the requested amounts do not follow a normal distribution, which is to be expected given the broad range of disasters. We may need to remove outliers in the analysis, but for now, we'll leave them to capture the true nature of disaster declarations. The largest outlier is likely Hurricane Maria. The log transformation does a better job of showing the distribution of the data, so we may need to consider an exponential model in the analysis. The transformed and untransformed figures suggest there are significant tails in the data.

@@

## Outcome of Interest: Obligated FEMA Funds
The alternative outcome of interest is the funding obligated from FEMA. While this may be the same as `ReqAmt` (especially for large-scale disasters), FEMA doesn't always pay the amount the states request.
```{r obligated funds}
#summary statistics
base::summary(EDAdata_outliers$OblAmt)

#density plot
ggplot2::ggplot(data = EDAdata_outliers, aes(x = OblAmt)) +
  geom_density()

#try a log transformation out of curiosity 
ggplot2::ggplot(data = EDAdata_outliers, aes(x = log(OblAmt))) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata_outliers, aes(y = OblAmt)) +
  geom_boxplot()

#crop to $1M to get a better idea
ggplot2::ggplot(data = EDAdata_outliers, aes(y = OblAmt)) +
  geom_boxplot()

#log transformed boxplot
ggplot2::ggplot(data = EDAdata_outliers, aes(y = log(OblAmt))) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata_outliers$OblAmt)
```
It appears that obligated funds and requested funds have similar distributions and characteristics. The same commentary about the requested funds analysis applies here as well.

Before moving on, I want to create a plot of Requested vs obligated Funds to understand the relationship between the two outcomes:
```{r}
#including all points
funds_comp <- EDAdata_outliers %>%
                ggplot2::ggplot(aes(x = log(ReqAmt), y = log(OblAmt))) +
                         geom_smooth(color = "#440154",
                                     size = 2) +
                         geom_point(color = "#1f9e89",
                                    alpha = 0.5) +
                         scale_y_continuous(expand = c(0,0),
                                            limits = c(0, NA)) +
                         scale_x_continuous(expand = c(0,0),
                                            limits = c(0, NA)) +
                         labs(x = "log(Requested Funding)",
                              y = "log(Obligated Funding)")

funds_comp

#save file
funds_comp_fig_file = here("results","funds_comp.png")
ggsave(filename = funds_comp_fig_file, plot = funds_comp)
```

I also want a density plot of both log-transformed outcomes on one plot, so that can go here as well.
```{r}
#convert to long format in order to be able to plot both outcomes with same variable
data_long <- gather(EDAdata_outliers, FundType, funds, ReqAmt:OblAmt, factor_key=TRUE)
data_long

#change labels
levels(data_long$FundType) <- c("Requested Amount", "Obligated Amount")

#now usual ggplot2 density plot
funds_density <- ggplot2::ggplot(data = data_long, aes(x = log(funds), color = FundType, fill = FundType)) +
                          geom_density(alpha = 0.4) +
                          facet_wrap(~ FundType) +
                          scale_fill_manual(values = c("#6ece58", "#26828e")) +
                          scale_color_manual(values = c("#6ece58", "#26828e")) +
                          labs(x = "Log(Funding)",
                               y = "Density") +
                          theme(legend.position="none")

funds_density

#save file
funds_density_fig_file = here("results","funds_density.png")
ggsave(filename = funds_density_fig_file, plot = funds_density)

```

@@

## Predictor: State
```{r}
#frequency table since it's categorical
#hide NAs because there are no NAs
summarytools::freq(EDAdata_outliers$state, report.nas = FALSE)

#examine graphical relationship with outcomes
#include a jitter function to have a better idea of number of measurements and distribution
#filter top 5 states
#start with obligated amount
obl_fig_state <- EDAdata_outliers %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dplyr::dense_rank(-n) < 5) %>%
                    ggplot2::ggplot(aes(x = state, y = OblAmt)) +
                      geom_boxplot(color = "#440154") +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      labs(x = "State or Territory",
                           y = "Obligated FEMA Funds")
obl_fig_state

#repeat for requested funds
req_fig_state <- EDAdata_outliers %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
                    ggplot2::ggplot(aes(x = state, y = ReqAmt)) +
                      geom_boxplot(color = "#440154") +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      labs(x = " " ,
                           y = "Requested FEMA Funds")
req_fig_state

#combine into one plot
plot_states <- cowplot::plot_grid(req_fig_state, 
                                     obl_fig_state,
                                     ncol = 1,
                                     align = "v",
                                     rel_heights = c(0.5,0.5))
plot_states

title_state <- cowplot::ggdraw() + 
                    draw_label(
                      "Requested vs Obligated FEMA Funds For Five Most Disaster Prone States",
                      fontface = 'bold',
                      x = 0,
                      hjust = 0) +
                    theme(plot.margin = margin(0, 0, 0, 7))

plot_states_funds <- cowplot::plot_grid(
                              title_state, plot_states,
                              ncol = 1,
                              # rel_heights values control vertical title margins
                              rel_heights = c(0.1, 1))
plot_states_funds

#save file
states_fig_file = here("results","states-funds.png")
ggsave(filename = states_fig_file, plot = plot_states_funds)

#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
decs_state_5 <- EDAdata_outliers %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dense_rank(-n) < 6) %>%
                    gtsummary::select(state, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = state,
                      label = list(state ~ "State or Territory",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**State**") %>%
                    gtsummary::modify_caption("**Table 4. FEMA Funding for 5 Most Disaster-Prone States 2012 - 2021**") %>%
                    gtsummary::bold_labels()
decs_state_5

@@

## Predictor: Incident Year
```{r}
#summary statistics since it's continuous
base::summary(EDAdata_outliers$IncidentYear)

@@

#density plot
ggplot2::ggplot(data = EDAdata_outliers, aes(x = IncidentYear)) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata_outliers, aes(y = IncidentYear)) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata_outliers$IncidentYear)

#histogram distribution
year_hist <- EDAdata_outliers %>% 
               ggplot2::ggplot(aes(x=IncidentYear)) + 
                        geom_bar(fill="#482878") + 
                        scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
                        scale_y_continuous(expand = c(0, 0), limits = c(0, 125)) +
                        geom_text(stat='count', aes(label=..count..), vjust=-1) +
                        labs(x = "Incident Year",
                             y = "Number of Declarations",
                             title = "Disaster Declarations per Year (2012 - 2021)")
year_hist

@@

#examine graphical relationship with outcomes of interest
#look at total cost per incident year
req_fig_year <- EDAdata_outliers %>%
                    ggplot2::ggplot(aes(x = IncidentYear)) +
                      geom_bar(aes(y = ReqAmt),
                               stat = "identity",
                               fill = "#6ece58") +
                      scale_y_continuous(expand = c(0, 0), 
                                         labels = scales::dollar_format()) +
                      scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
                      labs(x = "Incident Year",
                           y = "Funds ($)",
                           subtitle = "Requested")
req_fig_year

#repeat for obligated funds
obl_fig_year <- EDAdata_outliers %>%
                    ggplot2::ggplot(aes(x = IncidentYear)) +
                      geom_bar(aes(y = OblAmt),
                               stat = "identity",
                               fill = "#26828e") +
                      scale_y_continuous(expand = c(0, 0),
                                         labels = scales::dollar_format()) +
                      scale_x_continuous(breaks = seq(2012, 2021, by = 1)) +
                      labs(x = "Incident Year",
                           y = "Funds ($)",
                           subtitle = "Obligated")
obl_fig_year

@@

title1 <- cowplot::ggdraw() + 
                    draw_label(
                      "Requested vs Obligated FEMA Funds Per Year",
                      fontface = 'bold',
                      x = 0,
                      hjust = 0) +
                    theme( plot.margin = margin(0, 0, 0, 7))

@@

#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
decs_year_3 <- EDAdata_outliers %>%
                    dplyr::add_count(IncidentYear) %>%
                    dplyr::filter(dense_rank(-n) < 4) %>%
                    gtsummary::select(IncidentYear, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = IncidentYear,
                      label = list(IncidentYear ~ "Year",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Year**") %>%
                    gtsummary::modify_caption("**Table 2. FEMA Funding for the 3 Costliest Years (2012 - 2021)**") %>%
                    gtsummary::bold_labels()
decs_year_3

@@

## Incident Month
```{r}
#frequency table since it's categorical
#hide NAs because there are no NAs
summarytools::freq(EDAdata_outliers$IncidentMonth, report.nas = FALSE)

#examine graphical relationship with outcomes
#include a jitter function to have a better idea of number of measurements and distribution
#filter top 5 states
#start with obligated amount
obl_fig_month <- EDAdata_outliers %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dplyr::dense_rank(-n) < 5) %>%
                    ggplot2::ggplot(aes(x = IncidentMonth, y = OblAmt)) +
                      geom_boxplot(color = "#440154") +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      labs(x = "Incident Month",
                           y = "Funds ($)",
                           subtitle = "Obligated")
obl_fig_month

#repeat for requested funds
req_fig_month <- EDAdata_outliers %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dplyr::dense_rank(-n) < 6) %>%
                    ggplot2::ggplot(aes(x = IncidentMonth, y = ReqAmt)) +
                      geom_boxplot(color = "#440154") +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      labs(x = " " ,
                           y = "Funds ($)",
                           subtitle = "Requested")
req_fig_month

#combine into one plot
plot_month <- cowplot::plot_grid(req_fig_month, 
                                     obl_fig_month,
                                     ncol = 1,
                                     align = "v",
                                     rel_heights = c(0.5,0.5))
plot_month

title_month <- cowplot::ggdraw() + 
                    draw_label(
                      "Requested vs Obligated FEMA Funds By Month",
                      fontface = 'bold',
                      x = 0,
                      hjust = 0) +
                    theme(plot.margin = margin(0, 0, 0, 7))

plot_month_funds <- cowplot::plot_grid(
                              title_month, plot_month,
                              ncol = 1,
                              # rel_heights values control vertical title margins
                              rel_heights = c(0.1, 1))
plot_month_funds

#save file
month_fig_file = here("results","month-funds.png")
ggsave(filename = month_fig_file, plot = plot_month_funds)

#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
decs_month_5 <- EDAdata_outliers %>%
                    dplyr::add_count(IncidentMonth) %>%
                    dplyr::filter(dense_rank(-n) < 6) %>%
                    gtsummary::select(IncidentMonth, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = IncidentMonth,
                      label = list(IncidentMonth ~ "Incident Month",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**Incident Month**") %>%
                    gtsummary::modify_caption("**Table 4. FEMA Funding by Month 2012 - 2021**") %>%
                    gtsummary::bold_labels()
decs_month_5

#save file
decs_month_file = here("results", "decs-month-5.Rds")
saveRDS(decs_month_5, file = decs_month_file)
```

@@

## Predictor: Declaration Type
```{r}
#frequency table since it's categorical
#hide NAs because there are no NAs
summarytools::freq(EDAdata_outliers$declarationType, report.nas = FALSE)

#examine graphical relationship with outcomes
#include a jitter function to have a better idea of number of measurements and distribution
#start with obligated amount
obl_fig_DT <- EDAdata_outliers %>%
                    ggplot2::ggplot(aes(x = declarationType, y = OblAmt)) +
                      geom_boxplot() +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#26828e") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      scale_x_discrete(labels = c("Major Disaster Declaration", "Emergency Declaration")) + 
                      labs(x = "Declaration Type",
                           y = "Funds ($)",
                           subtitle = "Obligated")
obl_fig_DT

#repeat for requested funds
req_fig_DT <- EDAdata_outliers %>%
                    ggplot2::ggplot(aes(x = declarationType, y = ReqAmt)) +
                      geom_boxplot() +
                      geom_jitter(alpha = 0.5, width = 0.2, height = 0.2, color = "#6ece58") +
                      scale_y_continuous(labels = scales::dollar_format()) +
                      scale_x_discrete(labels = c("Major Disaster Declaration", "Emergency Declaration")) + 
                      labs(x = " ",
                           y = "Funds ($)",
                           subtitle = "Requested")
req_fig_DT

#combine into one plot
plot_DT <- cowplot::plot_grid(req_fig_DT, 
                                     obl_fig_DT,
                                     ncol = 1,
                                     align = "v",
                                     rel_heights = c(0.5,0.5))
plot_DT

title_DT <- cowplot::ggdraw() + 
                    draw_label(
                      "Requested vs Obligated FEMA Funds By Declaration Type",
                      fontface = 'bold',
                      x = 0,
                      hjust = 0) +
                    theme(plot.margin = margin(0, 0, 0, 7))

plot_DT_funds <- cowplot::plot_grid(
                              title_DT, plot_DT,
                              ncol = 1,
                              # rel_heights values control vertical title margins
                              rel_heights = c(0.1, 1))
plot_DT_funds

#save file
DT_fig_file = here("results","DT-funds.png")
ggsave(filename = DT_fig_file, plot = plot_DT_funds)
